\change{The effect of the COVID-19 pandemic on arXiv pre-prints in maths, physics, computer science, and quantitative biology}

### 2022-10-31 article/ms.tex
co-author comments

%%%%%%%%%%% Defining Enunciations  %%%%%%%%%%%
\newcommand{\arxiv}{arXiv}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[normalem]{ulem}
\newcommand{\del}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\add}[1]{\textcolor{blue}{#1}}
\newcommand{\todo}[1]{\textcolor{green}{#1}}
\newcommand{\change}[1]{\textbf{#1}}

@@

\author{%%%% Author details
A. R. Casey$^{1,2}$,  I. Mandel$^{1,3,4}$ and P. K. Ray$^{5}$}

@@

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics \change{in science and technology specialisations} are posting more \change{manuscripts} on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in general physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in quantitative biology where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians). 
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\noindent Peer-reviewed publications are the most common measure of productivity in \change{many specialties of science and technology}. The \arxiv\cite{Ginsparg:2011} ({https://arxiv.org}) is a service for the distribution of research manuscripts (pre-prints) prior to their appearance in a journal. The presence of a pre-print on \arxiv\ does not ensure that the contents have passed peer-review, but most material on \arxiv\ eventually goes through peer-review because it is now standard in many research specialties to post to \arxiv\ either during or after the peer-review process \cite{Lariviere:2014}. 
\change{The fraction of peer-reviewed publications that are also posted to the \arxiv\ increases each year\cite{Davis:2007} but not all academic outputs (peer reviewed publications or otherwise) will appear on the \arxiv. Fundamental reports, findings from inquiries or research groups, and many other texts that are only published in peer-reviewed publications will never appear on the \arxiv. In the specialisations we investigate here (physics, mathematics, astronomy, computer science, and other STEM-related fields), the fraction of peer-reviewed literature that also appears on \arxiv\ is estimated to vary from 18.5\% in mathematics to nearly 100\% in high energy physics\cite{Davis:2007,GentilBeccot:2009}. Although the fraction may vary between discplines, within a given specialisation the number of pre-prints provides a proxy measure of the number of peer-reviewed publications, which we can use to examine changes in each specialty as a function of time.}

Here we make quantitative comparisons on the number of pre-prints posted to \arxiv\ before and during the pandemic. It is important to recognise that the pandemic has affected both the general population\cite{Nicola:2020,Chu:2020,IbnMohammed:2021} and scientific community\cite{Viglione:2020,Gewen:2020,King:2021} in complex and unequal ways. Scientists have faced substantial challenges driven by a large number of factors including: generational inequality, carer responsibilities, modified work environments, and other personal and professional circumstances. Our analysis does not directly consider these important factors, nor does it account for groups or individuals disproportionately affected by the pandemic. We have instead adopted a broad perspective and present new quantitative results for entire scientific fields. The aim is to provide a “large-scale” view of scientific productivity which will serve as a foundation for future studies on the detailed impacts of COVID-19 on academic work.

@@

The field of quantitative biology (\texttt{q-bio}) showed the largest increase in pre-prints in 2020 above what is expected by our model, before returning to pre-pandemic levels in mid-2021. % todo: 
There were {2,790} quantitative biology pre-prints in 2020, {50\%} above the previous year, representing a $+8\sigma$ deviation from our model predictions (predicted 1,914 $\pm$ 109; 876 fewer than observed). This increase is explainable by an increase in COVID-19 related research, as there were 864 quantitative biology pre-prints in 2020 with pandemic-related terms in their title or abstract (see Methods), and just 39 in the decade prior. Indeed, in {April 2020} nearly 60\% of the \arxiv\ pre-prints in quantitative biology were related to the pandemic (Figure~\ref{fig:2}). Pandemic-related pre-prints also appeared in other fields (computer science, physics, statistics, and quantitative finance, all peaking around {April 2020}), but these papers constituted fewer than 10\% of the total pre-prints in these fields. Approximately 10--15\% of current pre-prints in quantitative biology are related to the ongoing pandemic. This represents a strong shift in research focus, but this fraction is declining with time. % todo:

\begin{figure}
 \includegraphics[width=\linewidth]{pandemic-related-preprints}
\caption{Pre-prints related to the COVID-19 pandemic peaked in April 2020, accounting for nearly 60\% the quantitative biology (\texttt{q-bio}) pre-prints in that month. Shown are five other research fields with the highest number of pandemic-related pre-prints in 2020. Monthly pre-prints are shown on the left, and the percent of pandemic-related pre-prints in that research field is shown on the right.}
\label{fig:2}
\end{figure}

@@

An increase in quantitative biology\footnote{Quantitative biology is only a sub-field of biology, and most biology pre-prints are posted to bioRxiv ({https://biorxiv.org}).} research during a global pandemic is unsurprising. However, the question arises if this increase was driven by existing researchers in the field, or by authors from other fields entering quantitative biology due to the pandemic. In 2020 there was a peak in the number of new authors appearing for the first time in the quantitative biology literature (Figure~\ref{fig:3}) while the number of new authors in all other fields remained steady. The increase in pre-prints (and new authors) is driven by small (1-4) groups of authors, where many had never posted pre-prints to quantitative biology before.

@@

One quarter of the excess pre-prints posted to quantitative biology during 2020 were both related to the pandemic and written by authors who had never appeared in the quantitative biology literature (234 of the 876 excess pre-prints in \texttt{q-bio}). A careful examination \change{(see Methods)} reveals that about half of these (45\%; 107/234) were written by mathematicians or physicists, with another 26\% (61/234) led by engineers, computer scientists, or economists. These pre-prints tend to focus on modelling the COVID-19 outbreak using public data sets, rather than quantitative biology research that requires more expert domain knowledge. About 3\% (8/234) were led by established biologists who have not used the \arxiv\ before and are now doing so presumably to help ensure that their COVID-19 research is more widely available. This fraction is likely higher among pre-prints where at least one author has appeared before in the quantitative biology literature. This represents a genuine increase in new researchers using the \arxiv, and a boost to making quantitative biology research more widely accessible.

@@

\change{The COVID-19 pandemic appears to have had minimal impact on collective research productivity {in the science and technology specialisations we examined}, {and that is before accounting for the tremendous work in scientific communication and contributions to public policy throughout the pandemic.} These findings are at odds with similar studies in fields that have larger research overlap with the SARS-CoV-2 virus. For example, publications in life sciences saw a 6.5 fold increase in subject headings related to COVID-19\cite{Riccaboni}. The effects of the pandemic clearly impact research specialisations differently, and we note some} significant changes within specific research areas. The largest persistent decline in pre-prints has occurred in {general physics (arXiv primary category \texttt{physics}), largely due to a decline in  pre-prints posted to computational and applied physics areas within the \texttt{physics} primary category. As of mid-2022, this steady decline continues.} Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020.

In this study we have only focussed on the quantity of \change{pre-prints} as a measure for productivity, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

\bibitem{Lariviere:2014}
{Larivi{\'e}re, V., et al.,} {arXiv E-prints and the journal of record: An analysis of roles and relationships}, \emph{Journal of the Association for Information Science and Technology}, \textbf{65}, 6, doi:10.1002/asi.23044 \newblock (2014).

>>>>

%%%%%%%%%%% Defining Enunciations  %%%%%%%%%%%
\newcommand{\arxiv}{arXiv}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[normalem]{ulem}
\newcommand{\del}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\add}[1]{\textcolor{blue}{#1}}
\newcommand{\todo}[1]{\textcolor{green}{#1}}
\newcommand{\change}[1]{\textcolor{blue}{\textbf{#1}}}

@@

\author{%%%% Author details
{A. Casey$^{1,2}$,  I. Mandel$^{1,3,4}$ and P. K. Ray$^{5}$}}

@@

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics \change{in science and technology specialisations} are posting more \change{manuscripts} on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in general physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in \change{the \arxiv\ category of} quantitative biology, where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians).
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\change{The \arxiv\ pre-print service\cite{Ginsparg:2011} ({https://arxiv.org}) has become an important platform for disseminating research in a growing number of scientific fields, most notably in physics, mathematics, and computer science. However, not all academic outputs will appear on the \arxiv: fundamental reports, findings from inquiries or research groups, and many other texts that are only published in peer-reviewed publications will never appear on the \arxiv, or any other pre-print server. In the specialisations we investigate (physics, mathematics, astronomy, computer science, and other STEM-related fields), the fraction of peer-reviewed literature that also appears on the \arxiv\ has been estimated to vary from 18.5\% in mathematics to nearly 100\% in high energy physics\cite{Davis:2007,GentilBeccot:2009}.}

 Here we make quantitative comparisons on the number of pre-prints posted to \arxiv\ before and during the pandemic. It is important to recognise that the pandemic has affected both the general population\cite{Nicola:2020,Chu:2020,IbnMohammed:2021} and scientific community\cite{Viglione:2020,Gewen:2020,King:2021} in complex and unequal ways. Scientists have faced substantial challenges driven by a large number of factors including: generational inequality, carer responsibilities, modified work environments, and other personal and professional circumstances. Our analysis does not directly consider these important factors, nor does it account for groups or individuals disproportionately affected by the pandemic. We have instead adopted a broad perspective and present new quantitative results for entire scientific fields. The aim is to provide a “large-scale” view of \change{one aspect of} scientific productivity which will serve as a foundation for future studies on the detailed impacts of COVID-19 on academic work.

@@

The \change{\arxiv\ category} of quantitative biology (\texttt{q-bio}) showed the largest increase in pre-prints in 2020 above what is expected by our model, before returning to pre-pandemic levels in mid-2021. % todo: 
There were {2,790} quantitative biology pre-prints in 2020, {50\%} above the previous year, representing a $+8\sigma$ deviation from our model predictions (predicted 1,914 $\pm$ 109; 876 fewer than observed). This increase is explainable by an increase in COVID-19 related research, as there were 864 \change{pre-prints in the quantitative biology \arxiv\ category} in 2020 with pandemic-related terms in their title or abstract (see Methods), and just 39 in the decade prior. Indeed, in {April 2020} nearly 60\% of the \change{pre-prints in the \arxiv\ category of quantitative biology} were related to the pandemic (Figure~\ref{fig:2}). Pandemic-related pre-prints also appeared in other fields (computer science, physics, statistics, and quantitative finance, all peaking around {April 2020}), but these papers constituted fewer than 10\% of the total pre-prints in these fields. Approximately 10--15\% of current pre-prints in \change{the \arxiv\ category of} quantitative biology are related to the ongoing pandemic. This represents a strong shift in research focus, but this fraction is declining with time. % todo:

\begin{figure}
 \includegraphics[width=\linewidth]{pandemic-related-preprints}
\caption{Pre-prints related to the COVID-19 pandemic peaked in April 2020, accounting for nearly 60\% the quantitative biology (\texttt{q-bio}) pre-prints in that month. Shown are five other categories with the highest number of pandemic-related pre-prints in 2020. Monthly pre-prints are shown on the left, and the percent of pandemic-related pre-prints in that research field is shown on the right.}
\label{fig:2}
\end{figure}

@@

An increase in quantitative biology\footnote{Quantitative biology is only a sub-field of biology, and most biology pre-prints are posted to bioRxiv ({https://biorxiv.org}).} research during a global pandemic is unsurprising. However, the question arises if this increase was driven by existing researchers in the field, or by authors from other fields \change{posting to the quantitative biology category} due to the pandemic. In 2020 there was a peak in the number of new authors appearing for the first time in the quantitative biology \change{\arxiv\ category} (Figure~\ref{fig:3}) while the number of new authors in all other fields remained steady. The increase in pre-prints (and new authors) is driven by small (1-4) groups of authors, where many had never posted pre-prints to \change{the \arxiv\ category of quantitative biology} before.

@@

One quarter of the excess pre-prints posted to \change{the quantitative biology \arxiv\ category} during 2020 were both related to the pandemic and written by authors who had never appeared in \change{that \arxiv\ category before} (234 of the 876 excess pre-prints in \texttt{q-bio}). A careful examination \change{(see Methods)} reveals that about half of these (45\%; 107/234) were written by mathematicians or physicists, with another 26\% (61/234) led by engineers, computer scientists, or economists. These pre-prints tend to focus on modelling the COVID-19 outbreak using public data sets, rather than quantitative biology research that requires more expert domain knowledge. About 3\% (8/234) were led by established biologists who have not used the \arxiv\ before. This fraction is likely higher among pre-prints where at least one author has appeared before in the quantitative biology \change{\arxiv\ category}. This represents a genuine increase in new researchers using the \arxiv, and a boost to making quantitative biology research more widely accessible.

@@

The COVID-19 pandemic appears to have had minimal impact on collective research productivity \change{in the science and technology specialisations we examined}, \change{and that is before accounting for the tremendous work in scientific communication and contributions to public policy throughout the pandemic.} \change{These findings are at odds with similar studies in fields that have larger research overlap with the SARS-CoV-2 virus. For example, publications in life sciences saw a 6.5 fold increase in subject headings related to COVID-19\cite{Riccaboni}. The effects of the pandemic clearly impact research specialisations differently, and we note some significant changes within specific research areas. The largest persistent decline in pre-prints has occurred in general physics (arXiv primary category \texttt{physics}), largely due to a decline in  pre-prints posted to computational and applied physics areas within the \texttt{physics} primary category. As of mid-2022, this steady decline continues. Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020.}

In this study we have only focussed on the quantity of \change{pre-prints}, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

%\bibitem{Lariviere:2014}
%{Larivi{\'e}re, V., et al.,} {arXiv E-prints and the journal of record: An analysis of roles and relationships}, \emph{Journal of the Association for Information Science and Technology}, \textbf{65}, 6, doi:10.1002/asi.23044 \newblock (2014).
